\documentclass[a4paper,11pt,oneside]{article}
\usepackage{url}
\usepackage{sectsty}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{fontawesome}
\usepackage[usenames,dvipsnames]{color}
\usepackage{hyperref}
\usepackage[normalem]{ulem}

\hypersetup{
    bookmarks=false,        % show bookmarks bar?
    unicode=true,           % non-Latin characters in Acrobat’s bookmarks
    pdfnewwindow=true,      % links in new window
    colorlinks=true,        % false: boxed links; true: colored links
    linkcolor=MidnightBlue, % color of internal links
    citecolor=green,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=MidnightBlue   % color of external links
}

\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\allsectionsfont{\sffamily}

%\newcommand{\CC}[1]{\textsf{\small #1}}% to quote from Computing Curricula 2001
\newcommand{\CC}[1]{#1}

\newcommand{\Cincluded}{{\small\cmark}}
\newcommand{\Cdefine}{{\small\cmark\faFileTextO}}
\newcommand{\Ccodeonly}{{\small\cmark\faFileText}}
\newcommand{\Cnofocus}{{\small\faQuestion}}
\newcommand{\Cexmaybe}{{\small\xmark\faQuestionCircle}}
\newcommand{\Cexcluded}{{\small\xmark}}

\newcommand{\Iincluded}{\item[\hbox to 1.8em{\Cincluded\hfill}]}
\newcommand{\Idefine}{\item[\hbox to 1.8em{\Cdefine\hfill}]}
\newcommand{\Icodeonly}{\item[\hbox to 1.8em{\Ccodeonly\hfill}]}
\newcommand{\Inofocus}{\item[\hbox to 1.8em{\Cnofocus\hfill}]}
\newcommand{\Iexmaybe}{\item[\hbox to 1.8em{\Cexmaybe\hfill}]}
\newcommand{\Iexcluded}{\item[\hbox to 1.8em{\Cexcluded\hfill}]}

\newcommand{\DRAFT}{ -- Version 2019}
\pagestyle{myheadings}
\textwidth 126.34 mm
\textheight 189.55 mm
\parindent 20 pt
\parskip 0 pt
\markboth{\sc The IOI Syllabus\DRAFT}{\sc The IOI Syllabus\DRAFT}

\newenvironment{myitemize}{\begin{quote}\begin{itemize}\itemsep 0pt}{\end{itemize}\end{quote}}

\newcommand{\remove}[1]{\sout{#1}}
\newcommand{\new}[1]{{\bf \color{red}{#1}}}
\newcommand{\richard}[1]{}


\begin{document}
\title{\sf The International Olympiad in Informatics Syllabus}
\date{~}
\maketitle

\section{Version and status information}%{{{

This is the Syllabus version intended for IOI 2019.

\medskip

There have been no changes since the 2018 version.
%All changes since the 2017 version are shown \new{in red}. 
%Nothing new was added, some topics were removed or restricted.

\medskip

The Syllabus is an official document related to the IOI.
For each IOI, an up-to-date version of
the Syllabus is produced by the ISC, as described in the IOI Regulations, Statue 3.13.
%}}}

\section{Authors and Contact Information}%{{{

The original proposal of the IOI Syllabus was co-authored by
{Tom Verhoeff}\footnote{TU Eindhoven, The Netherlands, \url{t.verhoeff@tue.nl}},
{Gyula Horv\'ath}\footnote{University of Szeged, Hungary, \url{horvath@inf.u-szeged.hu}},
{Krzysztof Diks}\footnote{Warsaw University, Poland, \url{diks@mimuw.edu.pl}},
and
{Gordon Cormack}\footnote{University of Waterloo, Canada, \url{gvcormac@uwaterloo.ca}}.

Since 2007, the following people have maintained the syllabus and made significant contributions:
{Michal Fori\v{s}ek}\footnote{Comenius University, Slovakia, \url{forisek@dcs.fmph.uniba.sk}},
{Jakub \L{}\c{a}cki}\footnote{Warsaw University, Poland, \url{j.lacki@mimuw.edu.pl}},
and
{Richard Peng}\footnote{Georgia Tech, USA, \url{richard.peng@gmail.com}}.

The most recent batch of revisions to the Syllabus was made by the ISC between February and July 2016.

You are welcome to send any feedback on the Syllabus to
the current maintainer's e-mail address (\verb!forisek@dcs.fmph.uniba.sk!).

For people interested in contributing to the quality of the Syllabus, some additional 
background on the Syllabus and other miscellaneous information can be found at
\url{http://ksp.sk/~misof/ioi-syllabus/}.

\newpage
%}}}

\section{Introduction}\label{sec:intro}%{{{

During the years, the Syllabus has evolved. Currently it has the following purposes:

\begin{itemize}
\item
It specifies a small set of required prerequisite knowledge.
Below, this is given in the category ``Included, unlimited''
and to some extent also in ``Included, to be defined''.

\item
It serves as a set of guidelines that help decide whether a task is 
suitable for the International Olympiad in Informatics (IOI).
Based on this document, the International Scientific Committee (ISC)
evaluates the task proposals when selecting the competition tasks.

\item
As a consequence of the previous item, another purpose of the Syllabus 
is to help the organizers of national olympiads prepare
their students for the IOI.

\end{itemize}

The Syllabus aims to achieve these goals by providing a classification of topics and 
concepts from mathematics and computer science. 
More precisely, this Syllabus classifies each topic into one of six categories.
Ordered by topic suitability, these are:

\begin{myitemize}
\Iincluded Included, unlimited
\Idefine   Included, to be defined
\Icodeonly Included, not for task description
\Inofocus  Outside of focus
\Iexmaybe  Excluded, but open to discussion
\Iexcluded Explicitly excluded
\end{myitemize}

\noindent
In the next section we explain the scope of each category.

\newpage
%}}}

\section{Categories}\label{sec:categories}%{{{

This Syllabus classifies a selection of topics into six different ca\-tegories. 
Obviously, such a set of topics can never be exhaustive.
Instead, the list given in this Syllabus should serve as examples that map out the boundary.
Topics not explicitly mentioned in the Syllabus should be classified as follows:
\begin{itemize}
\itemsep -3pt
\item Anything that is a prerequisite of an Included topic is also Included.
\item Anything that is an extension of an Excluded topic or similar to an Excluded topic is also Excluded.
\item Anything else that is not mentioned in the Syllabus is considered Outside of focus.
\end{itemize}

Note that issues related to the usage of suitable terminology and notations in competition tasks
are beyond the scope of this document.\footnote{See
T. Verhoeff: \emph{Concepts, Terminology, and Notations for IOI Competition Tasks},
\url{http://scienceolympiads.org/ioi/sc/documents/terminology.pdf}}

If there is a particular topic for which you are not sure how it should
be classified, we invite you to submit a clarification request to the 
current Syllabus maintainer.

\bigskip

\noindent
Here are the definitions of the six possible classifications:

\begin{description}
\item[\Cincluded\ Included, unlimited]~\\
    Topics in this category are considered to be prerequisite knowledge.
    Contestants are expected to know them. These topics can appear
    in task descriptions without further clarification.

    Example: \emph{Integer} in~\S\ref{subsubsec:NG}

\item[\Cdefine\ Included, to be defined]~\\
    Contestants should know this topic,
    but when it appears in a task description,
    the statement should contain a sufficient definition.
    This category is usually applied in situations where a general
    concept that would be \Cincluded\ has many different ``flavors'' and a formal definition
    is needed to distinguish among those.

    Example: \emph{Directed graph} in~\S\ref{subsubsec:DS}~DS2

\item[\Ccodeonly\ Included, not for task description]~\\
    Topics that belong to this category should not appear in tasks 
    descriptions. However, developing solutions and understanding 
    model solutions may require the knowledge of these topics.

    Example: \emph{Asymptotic analysis of upper complexity bounds\/}
    in \S\ref{subsubsec:AL}~AL1

    Note: This is the main category that should be of interest when 
    preparing contestants for the IOI. 
    It should be noted that this set of topics
    contains a wide range of difficulties, starting from simple concepts and ending 
    with topics that can appear in problems that aim to distinguish among 
    the gold medallists. It is \textbf{not} expected that all contestants
    should know everything listed in this category.

\item[\Cnofocus\ Outside of focus]~\\
    Any topic that is not explicitly addressed by the Syllabus
    should be considered to belong to this category.

    Contestants are not expected to have knowledge of these topics.
    Most competition tasks will not be related to any topics 
    from this category.

    However, this does not prevent the inclusion of 
    a competition task that is related to a particular topic
    from this category. The ISC may wish to include such a competition 
    task in order to broaden the scope of the IOI.
    
    If such a task is considered for the IOI,
    the ISC will make sure that the task can reasonably be solved
    without prior knowledge of the particular topic, and that 
    the task can be stated
    in terms of \Cincluded\ and \Cdefine\ concepts in a precise, concise, 
    and clear way.

    Examples of such tasks being used at recent IOIs include:
    \begin{itemize}
    \itemsep -3pt
    \item Languages (a.k.a. Wikipedia) from IOI 2010 in Canada
    \item Odometer (a.k.a. robot with pebbles) from IOI 2012 in Italy
    \item Art class from IOI 2013 in Australia.
    \end{itemize}


\item[\Cexcluded\ Explicitly excluded]~\\
    Some of the harder algorithmic topics are explicitly marked as excluded.
    It is guaranteed that there will not be a competition
    task that \emph{requires} the contestants to know these areas.

    Furthermore, the tasks will be set with the goal that knowledge of 
    Excluded topics should not help in obtaining simpler solutions / solutions 
    worth more points.

    This category contains topics whose inclusion will result in
    problems that are unaccessible to a significant portion of IOI participants.
    This includes but is not limited to hard textbook algorithms and advanced
    areas in mathematics.

    Still, note that the Syllabus must not be interpreted to restrict in 
    any way the techniques that contestants are allowed to apply in solving 
    the competition tasks.
    
    Examples: \emph{Calculus\/} in~\S\ref{subsubsec:other-mathematics}

\item[\Cexmaybe\ Excluded, but open to discussion]~\\
    As the Syllabus is a living document, there can be cases when we consider
    bringing in some of the Excluded topics. Usually, the topics in question
    are natural extensions of Included topics, or ones where drawing an exact
    boundary is difficult. Should such topics appear, they will be temporarily
    classified as ``Excluded, but open to discussion'', and by doing so we encourage 
    all members of the IOI community to give us feedback on these topics.
\end{description}

\bigskip

\noindent
The rest of this document contains the classification of topics.
%}}}

\section {Mathematics}
\label{subsec:mathematics}

\subsection {Arithmetics and Geometry}%{{{
\label{subsubsec:NG}

    \begin{myitemize}
    \Iincluded Integers, operations (incl.\ exponentiation), comparison
    \Iincluded Basic properties of integers (sign, parity, divisibility)
    \Iincluded Basic modular arithmetic: addition, subtraction, \\ multiplication
    \Icodeonly Prime numbers
    \Iincluded Fractions, percentages
    \Iincluded Line, line segment, angle, triangle, rectangle, square, circle
    \Iincluded Point, vector, coordinates in the plane
    \Iincluded Polygon (vertex, side/edge, simple, convex, inside, area)
    \Idefine Euclidean distances
    \Icodeonly Pythagorean theorem
    \end{myitemize}

    %{Excluded, but open to discussion}
    \begin{myitemize}
    \Iexmaybe Additional topics from number theory.
    \end{myitemize}
    
    %{Explicitly excluded\/}:
    \begin{myitemize}
    \Iexcluded geometry in 3D or higher dimensional spaces
    \Iexcluded analyzing and increasing precision of floating-point \\ computations
    \Iexcluded modular division and inverse elements
    \Iexcluded complex numbers,
    \Iexcluded general conics (parabolas, hyperbolas, ellipses)
    \Iexcluded trigonometric functions
    \end{myitemize}

%}}}
\subsection {Discrete Structures (DS)}%{{{
\label{subsubsec:DS}

    \subsubsection*{DS1. Functions, relations, and sets}

        \begin{myitemize}
        \Idefine\CC{Functions (surjections, injections, inverses, composition)}
        \Idefine\CC{Relations (reflexivity, symmetry, transitivity, equivalence relations,
        total/linear order relations, lexicographic order)}
        \Idefine\CC{Sets (inclusion/exclusion, complements, Cartesian products, power sets)}
        \end{myitemize}

        %{Explicitly excluded\/}:
        \begin{myitemize}
        \Iexcluded{Cardinality and countability} (of infinite sets)
        \end{myitemize}
  
    \subsubsection*{DS2. Basic logic}

        \begin{myitemize}
        \Iincluded First-order logic
        \Iincluded\CC{Logical connectives} (incl.\ their basic properties)
        \Iincluded\CC{Truth tables}
        \Iincluded\CC{Universal and existential quantification} (Note: statements should avoid definitions with nested quantifiers whenever possible.)
        \Icodeonly\CC{Modus ponens and modus tollens}
        \end{myitemize}

        %{Out of focus\/}:
        \begin{myitemize}
        \Inofocus \CC{Normal forms}
        \end{myitemize}
        
        %{Explicitly excluded\/}:
        \begin{myitemize}
        \Iexcluded \CC{Validity}
        \Iexcluded \CC{Limitations of predicate logic}
        \end{myitemize}

    \subsubsection*{DS3. Proof techniques}

        \begin{myitemize}
        \Idefine\CC{Notions of implication, converse, inverse, contrapositive, negation, and contradiction}
        \Icodeonly\CC{Direct proofs, proofs by: counterexample, contraposition, contradiction}
        \Icodeonly\CC{Mathematical induction}
        \Icodeonly\CC{Strong induction} (also known as complete induction)
        \Iincluded\CC{Recursive mathematical definitions} (incl.\ mutually recursive definitions)
        \end{myitemize}

    \subsubsection*{DS4. Basics of counting}

        \begin{myitemize}
        \Iincluded\CC{Counting arguments (sum and product rule, arithmetic and geometric progressions, Fibonacci numbers)}
        \Idefine\CC{Permutations and combinations (basic definitions)}
        \Idefine Factorial function, binomial coefficients
        \Icodeonly\CC{Inclusion-exclusion principle}
        \Icodeonly\CC{Pigeonhole principle}
        \Icodeonly\CC{Pascal's identity}, \CC{Binomial theorem}
        \end{myitemize}

        %{Explicitly excluded\/}:
        \begin{myitemize}
        \Iexcluded Solving of recurrence relations
        \Iexcluded Burnside lemma
        \end{myitemize}
  
    \subsubsection*{DS5. Graphs and trees}

        \begin{myitemize}
        \Idefine\CC{Trees} and their basic properties, rooted trees
        \Idefine\CC{Undirected graphs} (degree, path, cycle, connectedness, Euler/Hamil\-ton path/cycle, handshaking lemma)
        \Idefine\CC{Directed graphs} (in-degree, out-degree, directed path/cycle, Euler/Hamilton path/cycle)
        \Idefine\CC{Spanning trees}
        \Idefine\CC{Traversal strategies}
        \Idefine `Decorated' graphs with edge/node labels, weights, colors
        \Idefine Multigraphs, graphs with self-loops
        \Idefine Bipartite graphs
        \Icodeonly Planar graphs
        \end{myitemize}

        %{Explicitly Excluded}
        \begin{myitemize}
        \Iexcluded Hypergraphs
        \Iexcluded Specific graph classes such as perfect graphs
        \Iexcluded Structural parameters such as treewidth and expansion
        \Iexcluded Planarity testing
        \Iexcluded Finding separators for planar graphs
        \end{myitemize}

    \subsubsection*{DS6. Discrete probability}
  
        \begin{quote}
        Applications where everything is finite (and thus arguments about probability can be easily
        turned into combinatorial arguments) are \Cnofocus, everything more complicated
        is \Cexcluded.
        \end{quote}

%}}}
\subsection {Other Areas in Mathematics}%{{{
\label{subsubsec:other-mathematics}

    %{Explicitly excluded\/}:
    \begin{myitemize}
    \Iexcluded Geometry in three or more dimensions.
    \Iexcluded  Linear algebra, including (but not limited to):
        \begin{myitemize}
        \item Matrix multiplication, exponentiation, \\
              inversion, and Gaussian elimination
        \item Fast Fourier transform
        \end{myitemize}
    \Iexcluded Calculus
    \Iexcluded Theory of combinatorial games, e.g., NIM game, Sprague-Grundy theory
    \Iexcluded Statistics
    \end{myitemize}

%}}}

\section {Computing Science}
\label{subsec:computing-science}

\subsection {Programming Fundamentals (PF)}%{{{
\label{subsubsec:PF}

    \subsubsection*{PF1. Fundamental programming constructs\rm (for abstract machines)}

    \begin{myitemize}
    \Iincluded\CC{Basic syntax and semantics of a higher-level language} (at least one of the specific languages available at an IOI, as announced in the \emph{Competition Rules\/} for that IOI)
    \Iincluded\CC{Variables, types, expressions, and assignment}
    \Iincluded\CC{Simple I/O}
    \Iincluded\CC{Conditional and iterative control structures}
    \Iincluded\CC{Functions and parameter passing}
    \Icodeonly\CC{Structured decomposition}
    \end{myitemize}

    \subsubsection*{PF2. Algorithms and problem-solving}

    \begin{myitemize}
    \Icodeonly\CC{Problem-solving strategies} (understand--plan--do--check,
        separation of concerns, generalization, specialization, case distinction, working backwards,
        etc.)\footnote{See G. Polya: \emph{How to Solve It: A New Aspect of Mathematical Method},
        Princeton Univ.\ Press, 1948}
    \Icodeonly\CC{The role of algorithms in the problem-solving process}
    \Icodeonly\CC{Implementation strategies for algorithms} (also see \S\ref{subsec:software-engineering}~SE1)
    \Icodeonly\CC{Debugging strategies} (also see \S\ref{subsec:software-engineering}~SE3)
    \Idefine\CC{The concept and properties of algorithms} (correctness, efficiency)
    \end{myitemize}

    \subsubsection*{PF3. Fundamental data structures}

    \begin{myitemize}
    \Iincluded\CC{Primitive types} (boolean, signed/unsigned integer, character)
    \Iincluded\CC{Arrays} (incl. multicolumn dimensional arrays)
    \Iincluded\CC{Strings and string processing}
    \Idefine\CC{Static and stack allocation} (elementary automatic memory management)
    \Idefine\CC{Linked structures} 
    \Idefine\CC{Implementation strategies for graphs and trees}
    \Idefine\CC{Strategies for choosing the right data structure}
    \Icodeonly\CC{Elementary use of real numbers in numerically stable tasks}
    \Icodeonly The floating-point representation of real numbers, the existence of precision issues.\footnote{Whenever possible, avoiding floating point computations completely is the preferred solution.}
    \Icodeonly \CC{Pointers and references}
    \end{myitemize}

    \begin{myitemize}
    \Inofocus \CC{Data representation in memory},
    \Inofocus \CC{Heap allocation},
    \Inofocus  \CC{Runtime storage management},
    \Inofocus Using fractions to perform exact calculations.
    \end{myitemize}
    
    \begin{myitemize}
    \Iexcluded Non-trivial calculations on floating point numbers, manipulating precision errors
    \end{myitemize}
   
    \begin{quote}
    Regarding floating point numbers, there are well-known reasons why they
    should be, in general, avoided at the IOI.\footnote{%
    See G. Horv\'ath and T. Verhoeff: \emph{Numerical Difficulties in Pre-University Education and
    Competitions}, Informatics in Education 2:21--38, 2003}
    However, the currently used interface removes some of those issues.
    In particular, it should now be safe to use floating point numbers
    in some types of tasks -- e.g., to compute some Euclidean distances and 
    return the smallest one.
    \end{quote}

    \subsubsection*{PF4. Recursion}

    \begin{myitemize}
    \Iincluded\CC{The concept of recursion}
    \Iincluded\CC{Recursive mathematical functions}
    \Iincluded\CC{Simple recursive procedures} (incl.\ mutual recursion)
    \Icodeonly\CC{Divide-and-conquer strategies}
    \Icodeonly\CC{Implementation of recursion}
    \Icodeonly\CC{Recursive backtracking}
    \end{myitemize}

    \subsubsection*{PF5. Event-driven programming}

    \begin{quote}
    Some competition tasks may involve a dialog with a reactive environment.
    Implementing such an interaction with the provided environment is \Cdefine.

    Everything not directly related to the implementation of reactive tasks is \Cnofocus.
    \end{quote}

% }}}

\subsection {Algorithms and Complexity (AL)}
\label{subsubsec:AL}

    \begin{quote}
    We quote from the IEEE-CS Curriculum:

    \sffamily\small
    Algorithms are fundamental to computer science and software engineering.
    The real-world performance of any software system depends only on two things:
    (1)~the algorithms chosen and
    (2)~the suitability and efficiency of the various layers of implementation.
    Good algorithm design is therefore crucial for the performance of all software systems.
    Moreover,
    the study of algorithms provides insight into the intrinsic nature of the problem
    as well as possible solution techniques independent of programming language,
    programming paradigm, computer hardware, or any other implementation aspect.
    \end{quote}

    \subsubsection*{AL1. Basic algorithmic analysis}%{{{

    \begin{myitemize}
    \Idefine Algorithm specification, precondition, postcondition, correctness, invariants
    \Icodeonly\CC{Asymptotic analysis of upper complexity bounds} (informally if possible)
    \Icodeonly\CC{Big O notation}
    \Icodeonly\CC{Standard complexity classes}: constant, logarithmic, linear, $\mathcal{O}(n \log n)$, quadratic, cubic, exponential, etc.
    \Icodeonly\CC{Time and space tradeoffs in algorithms}
    \Icodeonly Empirical performance measurements.
    \end{myitemize}

    \begin{myitemize}
    \Inofocus \CC{Identifying differences among best, average, and worst case behaviors},
    \Inofocus \CC{Little o, Omega, and Theta notation},
    \Inofocus Tuning parameters to reduce running time, memory consumption or other measures of performance
    \end{myitemize}
  
    \begin{myitemize}
    \Iexcluded Asymptotic analysis of average complexity bounds
    \Iexcluded Using recurrence relations to analyze recursive algorithms
    \end{myitemize}
  
    %}}}
    \subsubsection*{AL2. Algorithmic strategies}%{{{

    \begin{myitemize}
    \Icodeonly Simple loop design strategies
    \Icodeonly\CC{Brute-force algorithms} (exhaustive search)
    \Icodeonly\CC{Greedy algorithms}
    \Icodeonly\CC{Divide-and-conquer}
    \Icodeonly\CC{Backtracking} (recursive and non-recursive), \CC{Branch-and-bound}
    \Icodeonly\CC{Dynamic programming}\footnote{The IEEE-CS Curriculum puts this under AL8, but we believe it belongs here.} 
    \end{myitemize}
  
    \begin{myitemize}
    \Inofocus Heuristics
    \Inofocus Finding good features for machine learning tasks\footnote{E.g., finding a good way to classify images in the IOI 2013 Art class problem.}
    \Inofocus Discrete approximation algorithms
    \Inofocus Randomized algorithms.
    \end{myitemize}
  
    \begin{myitemize}
    \Iexcluded Clustering algorithms (e.g. $k$-means, $k$-nearest neighbor)
    \Iexcluded Minimizing multi-variate functions using numerical approaches.
    \end{myitemize}
  
    %}}}
    \subsubsection*{AL3a. Algorithms}%{{{

    \begin{myitemize}
    \Icodeonly Simple algorithms involving integers: radix conversion, Euclid's algorithm, primality test by $\mathcal{O}(\sqrt{n})$ trial division, Sieve of Eratosthenes, factorization (by trial division or a sieve), efficient exponentiation
    \Icodeonly Simple operations on arbitrary precision integers (addition, subtraction, simple multiplication)\footnote{The necessity to implement these operations should be obvious from the problem statement.}
    \Icodeonly Simple array manipulation (filling, shifting, rotating, reversal, resizing, minimum/maximum, prefix sums, histogram, bucket sort)
    \Icodeonly Simple string algorithms (e.g., naive substring search)
    \Icodeonly\CC{sequential} processing/search \CC{and binary search}
    \Icodeonly \CC{Quicksort} and Quickselect to find the $k$-th smallest element.
    \Icodeonly\CC{$\mathcal{O}(n \log n)$} worst-case \CC{sorting algorithms (heap sort, merge sort)}
    \Icodeonly Traversals of ordered trees (pre-, in-, and post-order)
    \Icodeonly\CC{Depth- and breadth-first traversals}
    \Icodeonly Applications of the depth-first traversal tree, such as topological ordering and Euler paths/cycles
    \Icodeonly Finding connected components and transitive closures.
    \Icodeonly Shortest-path algorithms (Dijkstra, Bellman-Ford, Floyd-Warshall)
    \Icodeonly Minimum spanning tree (Jarn\'\i k-Prim and Kruskal algorithms)
    \Icodeonly $O(VE)$ time algorithm for computing maximum bipartite matching.
    \Icodeonly Biconnectivity in undirected graphs (bridges, articulation points).
    \Icodeonly Connectivity in directed graphs (strongly connected components).
    \Icodeonly Basics of combinatorial game theory, winning and losing positions, minimax algorithm for optimal game playing
    \end{myitemize}

    \begin{myitemize}
    \Iexmaybe Maximum flow. Flow/cut duality theorem.
    \end{myitemize}

    \begin{myitemize}
    \Iexcluded Optimization problems that are easiest to analyze using matroid theory. Problems based on matroid intersecions (except for bipartite matching).
    \Iexcluded Lexicographical BFS, maximum adjacency search and their properties
    \end{myitemize}
 
    %}}}
    \subsubsection*{AL3b. Data structures}%{{{
  
    \begin{myitemize}
    \Idefine Stacks and queues
    \Icodeonly \CC{Representations of graphs (adjacency lists, adjacency matrix)}
    \Idefine Binary heap data structures
    \Icodeonly Representation of disjoint sets: the Union-Find data structure.
    \Idefine Statically balanced binary search trees. Instances of this include binary index trees (also known as Fenwick trees)
    and segment trees (also known as interval trees and tournament trees).\footnote{Not to be confused with similarly-named data structures used in computational geometry.}
    \Icodeonly {Balanced binary search trees}\footnote{Problems will not be designed to distinguish between
    the implementation of BBSTs, such as treaps, splay trees, AVL trees, or scapegoat trees}
    \Idefine Augmented binary search trees
    \Icodeonly $O(\log n)$ time algorithms for answering lowest common ancestor queries in a static rooted tree.\footnote{Once again, different implementations meeting this requirement will not be distinguished.}
    \Icodeonly Creating persistent data structures by path copying.%\footnote{Gains in memory using methods such as fat-node will not be distinguished}}
    \Icodeonly Nesting of data structures, such as having a sequence of sets.
    \Idefine Tries
    \end{myitemize}

    \begin{myitemize}
    \Iexmaybe String algorithms and data structures: there is evidence that the inter-reducibility between KMP, Rabin-Karp hashing, suffix arrays/tree, suffix automaton, and Aho-Corasick makes them difficult to separate.
    \Iexmaybe Heavy-light decomposition and separator structures for static trees.
    \Iexmaybe Data structures for dynamically changing trees and their use in graph algorithms.
    \end{myitemize}

    \begin{myitemize}
    \Iexcluded Complex heap variants such as binomial and Fibonacci heaps,
    \Iexcluded Using and implementing \CC{hash tables} (incl. strategies to resolve collisions)
    \Iexcluded Two-dimensional tree-like data structures (such as a 2D statically balanced binary tree or a treap of treaps) used for 2D queries.
    \Iexcluded Fat nodes and other more complicated ways of implementing persistent data structures.
    \end{myitemize}

    %}}}
    \subsubsection*{AL4. Distributed algorithms}%{{{

    \begin{quote}
    This entire section is \Cnofocus.
    \end{quote}

    %}}}
    \subsubsection*{AL5. Basic computability}%{{{

    \begin{quote}
    All topics related to computability are \Cexcluded.
    This includes the following:
    \CC{Tractable and intractable problems};
    \CC{Uncomputable functions};
    \CC{The halting problem};
    \CC{Implications of uncomputability}.
    
    However, see AL7 for basic computational models.
    \end{quote}

    %}}}
    \subsubsection*{AL6. The complexity classes P and NP}%{{{

    \begin{quote}
    Topics related to non-determinism, proofs of NP-hardness (reductions),
    and everything related is \Cexcluded.

    Note that this section only covers the results usually contained in
    undergraduate and graduate courses on formal languages and 
    computational complexity. The classification of these topics
    as \Cexcluded\ does not mean that an NP-hard 
    problem cannot appear at an IOI.
    \end{quote}

    %}}}
    \subsubsection*{AL7. Automata and grammars}%{{{

    \begin{myitemize}
    \Idefine Understanding a simple grammar in Backus-Naur form
    \end{myitemize}

    \begin{myitemize}
    \Inofocus Formal definition and properties of finite-state machines,
    \Inofocus Context-free grammars and related rewriting systems,
    \Inofocus Regular expressions
    \end{myitemize}
  
    \begin{myitemize}
    \Iexcluded Properties other than the fact that automata are graphs and that grammars have parse trees.
    \end{myitemize}
  
    %}}}
    \subsubsection*{AL8. Advanced algorithmic analysis}%{{{

    \begin{myitemize}
    \Icodeonly Amortized analysis.
    \end{myitemize}

    \begin{myitemize}
    \Inofocus Online algorithms
    \Inofocus Randomized algorithms
    \end{myitemize}

    \begin{myitemize}
    \Iexcluded Alpha-beta pruning
    \end{myitemize}
    
    %}}}
    \subsubsection*{AL9. Cryptographic algorithms}%{{{

    \begin{quote}
    This entire section is \Cnofocus.
    \end{quote}

    %}}}
    \subsubsection*{AL10. Geometric algorithms}%{{{

    \begin{quote}
    In general, the ISC has a strong preference towards problems that can be solved using integer
    arithmetics to avoid precision issues. This may include representing some computed values as 
    exact fractions, but extensive use of such fractions in calculations is discouraged.

    Additionally, if a problem uses two-dimensional objects, the ISC prefers problems in which such objects are rectilinear.
    \end{quote}

    \begin{myitemize}
    \Icodeonly Representing points, vectors, lines, line segments.
    \Icodeonly Checking for collinear points, parallel/orthogonal vectors and clockwise turns (for example, by using dot products and cross products).
    \Icodeonly Intersection of two lines.
    \Icodeonly Computing the area of a polygon from the coordinates of its vertices.\footnote{The recommended way of doing so is to use cross products or an equivalent formula. TODO url}
    \Icodeonly Checking whether a (general/convex) polygon contains a point.
    \Icodeonly Coordinate compression.
    \Icodeonly $\mathcal{O}(n\log{n})$ time algorithms for convex hull
    \Icodeonly Sweeping line method
    \end{myitemize}
  
    \begin{myitemize}
    \Iexcluded Point-line duality
    \Iexcluded Halfspace intersection, Voronoi diagrams, Delaunay triangulations.
    \Iexcluded Computing coordinates of circle intersections against lines and circles.
    \Iexcluded Linear programming in 3 or more dimensions and its geometric interpretations.
    \Iexcluded Center of mass of a 2D object.
    \Iexcluded Computing and representing the composition of geometric transformations if the knowledge of linear algebra gives an advantage.
    \end{myitemize}

    %}}}
    \subsubsection*{AL11. Parallel algorithms}%{{{

    \begin{quote}
    This entire section is \Cnofocus.
    \end{quote}

    %}}}

\subsection{Other Areas in Computing Science}%{{{
\label{subsubsec:other-areas}

    \begin{quote}
    Except for GV (specified below), all areas are \Cexcluded.

    \begin{description}
    \item[AR. Architecture and Organization] ~
    \item[OS. Operating Systems] ~
    \item[NC. Net-Centric Computing] (a.k.a. cloud computing)
    \item[PL. Programming Languages] ~ 
    \item[HC. Human-Computer Interaction] ~ 
    \item[GV. Graphics and Visual Computing] ~ 
    
    Basic aspects of processing graphical data are \Cnofocus,
    everything else (including the use of graphics libraries such as OpenGL)
    is \Cexcluded.

    \item[IS. Intelligent Systems] ~
    \item[IM. Information Management] ~
    \item[SP. Social and Professional Issues] ~
    \item[CN. Computational Science] ~ 
    \end{description}

    Notes: AR is about digital systems, assembly language, instruction pipelining, cache memories, etc.
    OS is about the \emph{design\/} of operating systems, not their usage.
    PL is about the \emph{analysis and design\/} of programming languages, not their usage.
    HC is about the \emph{design\/} of user interfaces.

    \emph{Usage} of the operating system, GUIs and programming languages is 
    covered in \S\ref{subsec:computer-literacy} and \S\ref{subsubsec:PF}.
    \end{quote}
%}}}

\section {Software Engineering (SE)}%{{{
\label{subsec:software-engineering}

    \begin{quote}
    We quote from~the IEEE-CS Curriculum:

    {\sffamily\footnotesize
    Software engineering is the discipline concerned with the application of theory,
    knowledge, and practice for effectively and efficiently building software systems
    that satisfy the requirements of users and customers.
    }

    In the IOI competition,
    the application of software engineering concerns the use of light-weight techniques
    for small, one-off, single-developer projects under time pressure.
    All included topics are~\Ccodeonly.
    \end{quote}

    \subsubsection*{SE1. Software design}

        \begin{myitemize}
        \Icodeonly\CC{Fundamental design concepts and principles}
        \Icodeonly\CC{Design patterns}
        \Icodeonly\CC{Structured design}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Transform an abstract algorithm into a concrete, efficient program expressed in one of the allowed programming languages, possibly using standard or competition-specific libraries.
        \item[--] Make their programs read data from and write data to text files according to a prescribed simple format
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Software architecture},
        \Iexcluded \CC{Design for reuse},
        \Iexcluded \CC{Object-Oriented analysis and design},
        \Iexcluded \CC{Component-level design}
        \end{myitemize}

    \subsubsection*{SE2. Using APIs}

        \begin{myitemize}
        \Icodeonly\CC{API (Application Programming Interface) programming}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Use competition-specific libraries according to the provided specification.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Programming by example},
        \Iexcluded \CC{Debugging in the API environment},
        \Iexcluded \CC{Class browsers and related tools},
        \Iexcluded \CC{Introduction to component-based computing}
        \end{myitemize}

    \subsubsection*{SE3. Software tools and environments}

        \begin{myitemize}
        \Icodeonly\CC{Programming environments}, incl.\ IDE (Integrated Development Environment)
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Write and edit program texts using one of the provided program editors.
        \item[--] Compile and execute their own programs.
        \item[--] Debug their own programs.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Testing tools},
        \Iexcluded \CC{Configuration management tools}
        \Iexcluded \CC{Requirements analysis and design modeling tools},
        \Iexcluded \CC{Tool integration mechanisms}
        \end{myitemize}

    \subsubsection*{SE4. Software processes}

        \begin{myitemize}
        \Icodeonly\CC{Software life-cycle and process models}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Understand the various phases in the solution development process and select appropriate approaches.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Process assessment models},
        \Iexcluded \CC{Software process metrics}
        \end{myitemize}

    \subsubsection*{SE5. Software requirements and specification}

        \begin{myitemize}
        \Icodeonly\CC{Functional and nonfunctional requirements}\\
        \Icodeonly\CC{Basic concepts of formal specification techniques}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Transform a precise natural-language description (with or without mathematical formalism) into a problem in terms of a computational model, including an understanding of the efficiency requirements.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Prototyping},
        \Iexcluded \CC{Requirements elicitation},
        \Iexcluded \CC{Requirements analysis modeling techniques}
        \end{myitemize}

    \subsubsection*{SE6. Software validation}

        \begin{myitemize}
        \Icodeonly\CC{Testing fundamentals, including test plan creation and test case generation}
        \Icodeonly\CC{Black-box and white-box testing techniques}
        \Icodeonly\CC{Unit, integration, validation, and system testing}
        \Icodeonly\CC{Inspections}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Apply techniques that maximize the the opportunity to detect common errors (e.g.\ through well-structured code, code review, built-in tests, test execution).
        \item[--] Test (parts of) their own programs.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Validation planning},
        \Iexcluded \CC{Object-oriented testing}
        \end{myitemize}

    \subsubsection*{SE7. Software evolution}
        
        \begin{myitemize}
        \Iexcluded \CC{Software maintenance},
        \Iexcluded \CC{Characteristics of maintainable software},
        \Iexcluded \CC{Re-engineering},
        \Iexcluded \CC{Legacy systems},
        \Iexcluded \CC{Software reuse}
        \end{myitemize}

    \subsubsection*{SE8. Software project management}

        \begin{myitemize}
        \Icodeonly\CC{Project scheduling} (especially time management)
        \Icodeonly\CC{Risk analysis}
        \Icodeonly\CC{Software configuration management}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Manage time spent on various activities.
        \item[--] Weigh risks when choosing between alternative approaches.
        \item[--] Keep track of various versions and their status while developing solutions.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded \CC{Software quality assurance},
        \Iexcluded \CC{Team management},
        \Iexcluded \CC{Software measurement and estimation techniques},
        \Iexcluded \CC{Project management tools}
        \end{myitemize}

    \subsubsection*{SE9. Component-based computing}

        \begin{quote}
        This entire section is \Cexcluded.
        \end{quote}

    \subsubsection*{SE10. Formal methods}

        \begin{myitemize}
        \Icodeonly\CC{Formal methods concepts} (notion of correctness proof, invariant)
        \Icodeonly\CC{Pre and post assertions}
        \end{myitemize}

        \begin{quote}
        In particular, contestants may be expected to
        \begin{itemize}
        \itemsep -3pt
        \item[--] Reason about the correctness and efficiency of algorithms and programs.
        \end{itemize}
        \end{quote}

        \begin{myitemize}
        \Iexcluded\CC{Formal verification},
        \Iexcluded\CC{Formal specification languages},
        \Iexcluded\CC{Executable and non-executable specifications}
        \end{myitemize}

    \subsubsection*{SE11. Software reliability}

        \begin{quote}
        This entire section is \Cexcluded.
        \end{quote}

    \subsubsection*{SE12. Specialized systems development}
  
        \begin{quote}
        This entire section is \Cexcluded.
        \end{quote}

%}}}

\section {Computer Literacy}%{{{
\label{subsec:computer-literacy}

\begin{quote}
The text of this section is \Ccodeonly.

\medskip

Contestants should know and understand the basic structure and operation of a computer
(CPU, memory, I/O).
They are expected to be able to use a standard computer with graphical user interface,
its operating system with supporting applications,
and the provided program development tools
for the purpose of solving the competition tasks.
In particular,
some skill in file management is helpful (creating folders, copying and moving files).

Details of these facilities will be stated in the \emph{Competition Rules\/}
of the particular IOI.
Typically, some services are available through a standard web browser.
Possibly, some competition-specific tools are made available,
with separate documentation.

It is often the case that a number of equivalent tools are made available.
The contestants are not expected to know all the features of all these tools.
They can make their own choice based on what they find most appropriate.

\medskip

The following topics are all \Cnofocus:
  Calculator,
  Word-processors,
  Spreadsheet applications,
  Database management systems,
  E-mail clients,
  Graphics tools (drawing, painting).
\end{quote}
%}}}

\end{document}
% vim: fdm=marker
